Experiments and Modeling of the Autoignition of Methyl Pentanoate at Low to Intermediate Temperatures and Elevated Pressures in a Rapid Compression Machine

### 2017-09-22 revision-1/revised-paper/methyl-valerate.tex
Revision 1 sent to Jackie

\bibliographystyle{elsarticle-num-CNF}
\usepackage{doi}
\usepackage[capitalize]{cleveref}
\begin{document}
%======================================================================
\begin{frontmatter}
\title{ Experiments and Modeling of the Autoignition of Methyl Valerate at Low to Intermediate Temperatures and Elevated Pressures in a Rapid Compression Machine }

@@

This work provides additional data for the autoignition of MV. Data is collected in a RCM under
engine relevant conditions spanning from \SIrange{15}{30}{\bar}, equivalence ratios \((\phi)\) from
\numrange{0.25}{2.0}, and temperatures from \SIrange{680}{1050}{\K}. The NTC region of MV is mapped
out to provide additional information on the fidelity of using MV as a biodiesel surrogate.

@@

The RCM is equipped with heaters to control the initial temperature of the mixture. After filling in
the components to the mixing tanks, the heaters are switched on and the system is allowed
\SI{1.5}{\hour} to come to steady state. The mixing tanks are also equipped with magnetic stir bars
so the reactants are well mixed for the duration of the experiments.

@@

The initial temperature is chosen such that the saturated vapor pressure \((P_{\text{sat}})\) of the
fuel at the initial temperature is at least twice the partial pressure of the fuel in the mixing
tank. The Antoine equation
%
\begin{equation}\label{eq:antoine}
    \logten{P_{\text{sat}}} = A - \frac{B}{T - C}
\end{equation}
%
is used to model the saturated vapor pressure of MV as a function of temperature \((T)\), where
\(A\), \(B\), and \(C\) are substance-specific coefficients. Coefficients for \cref{eq:antoine} are
given in the literature by~\citet{Ortega2003}, \citet{Camacho2007}, and~\citet{Stephenson1987}.
Unfortunately, the values of the coefficients are different among all three references. Therefore,
coefficients for use in \cref{eq:antoine} are determined in this work by least squares fitting of
the data of \citet{Ortega2003}, \citet{vanGenderen2002}, and \citet{Verevkin2008} using the
\verb|curve_fit()| function of SciPy~\cite{Jones2001} version 0.19.0. \Cref{fig:vapor-pressure}
shows that the coefficients fitted with this procedure give good agreement with the experimental
data; values for the coefficients computed in this work and in the literature works are given in
\cref{tab:antoine}. The data and code used to calculate the coefficients are provided in the
Supplementary Material.

\begin{center}
    \captionsetup{type=table}
    \caption{Antoine Equation coefficients computed in this work and from the
    literature. The $2\sigma$ confidence interval is estimated by taking the
    square root of the diagonals of the covariance matrix returned from
    \texttt{curve\_fit()}}
    \label{tab:antoine}
    \begin{tabular}{cccccc}
        \toprule
        & $A$ & $B$ & $C$ & $T_{\text{min}}$, \si{\K} & $T_{\text{max}}$, \si{\K} \\
        \midrule
        This Work & 6.4030 & 1528.69 & 52.881 & 274.9 & 417.18 \\
        $2\sigma$ Confidence Interval & 0.0919 & 53.47 & 4.934 & --- & --- \\
        \citet{Ortega2003} & 6.23175 & 1429.00 & 62.30 & 364.75 & 417.18 \\
        \citet{Camacho2007} & 5.9644 & 1281.06 & 75.94 & 281 & 547 \\
        \citet{Stephenson1987} & 6.62646 & 1658.4 & 42.09 & 297 & 411 \\
        \bottomrule
    \end{tabular}
\end{center}

@@

\begin{center}
    \captionsetup{type=figure}
    \includegraphics[width=\textwidth]{figures/pressure-traces.pdf}
    \caption{Selected pressure traces around the NTC region of ignition delay.
    a) \(\phi=1.0,\ P_C=\SI{30}{\bar}\), b) \(\phi=2.0,\ P_C=\SI{15}{\bar}\)}.
    \label{fig:pressure-traces}
\end{center}

@@

\begin{center}
    \captionsetup{type=figure}
    \resizebox{\textwidth}{!}{\input{figures/simulation-comparison.pgf}}
    \caption{Comparison of experimental and simulated results. a) \SI{15}{\bar}, b) \SI{30}{\bar}}.
    \label{fig:simulation-comparison}
\end{center}

At \SI{15}{\bar}, the experimental ignition delays are under-predicted by the \citet{Dievart2013}
model for the three equivalence ratios shown. For the \(\phi =\) \numlist{0.25;0.5} conditions, the
model appears to be predicting an NTC region of the ignition delays as the temperature decreases,
although such a trend is not observed in the data. However, at \(\phi = 2.0\), the model does not
predict the presence of an NTC region, although one is present in the experiments. Nonetheless, the
agreement seems to be improving as the temperature is decreased. Comparing the \citet{Dievart2013}
model to the stoichiometric data at \SI{30}{\bar}, we find a similar trend as the \(\phi=2.0,\
P_C=\SI{15}{\bar}\) data. The model does not predict the NTC region found experimentally for the
\(\phi = 1.0\) experiments, but the agreement improves as the temperature decreases. Interestingly,
two-stage ignition is predicted for all of the \(\phi=1.0\) and \(\phi=2.0\) data shown in
\cref{fig:simulation-comparison}. However, the first-stage ignition delays are
\SIrange{0.1}{0.5}{\ms} less than the overall ignition delays, and are not shown on
\cref{fig:simulation-comparison} because they are nearly indistinguishable from the overall ignition
delay. While the model of \citet{Dievart2013} over-predicts the first-stage ignition delay, it also
over-predicts the first-stage pressure rise, thereby leading to rapid overall ignition.

To elucidate the underlying reasons for the disagreement between the \citet{Dievart2013} model and
the data, we constructed an additional model using RMG (see
\cref{sec:reaction-mechanism-generator}). As can be seen in \cref{fig:simulation-comparison}a, the
agreement between the RMG model (dashed lines) and the experimental data is similar to the
\citet{Dievart2013} model for the \SI{15}{\bar}, \(\phi =\) \numlist{0.25; 0.5} data. Moreover, the
RMG model predicts a similar NTC region as temperature is decreasing. For the \SI{15}{\bar}, \(\phi =
2.0\) data, the RMG model tends to over-predict the low-temperature ignition delays, and does not
predict the NTC region found experimentally. As before, the trend at \SI{30}{\bar}, \(\phi=1.0\) is
similar to the \SI{15}{\bar}, \(\phi=2.0\) data; the RMG model over-predicts the low-temperature
ignition delays and does not predict the experimental NTC region. Finally, as in the
\citet{Dievart2013} model, two-stage ignition is predicted for all of the \(\phi=1.0\) and
\(\phi=2.0\) data shown in \cref{fig:simulation-comparison}. However, the first-stage ignition
delays are \SIrange{0.1}{0.5}{\ms} less than the overall ignition delays, and are not shown on
\cref{fig:simulation-comparison} because they are nearly indistinguishable from the overall ignition
delay. Again, the over-prediction of the first-stage ignition delay and pressure rise requires
further investigation and understanding of the complex low-temperature chemistry.

@@

This difference in heats of formation affects the pathways that consume the fuel. By conducting a
reaction pathway analysis to determine which radicals are formed from the breakdown of the fuel, we
can analyze the proportion of each radical formed as the fuel breaks down during the autoignition
process. The following analysis is conducted for a constant volume, adiabatic simulation with
initial temperature and pressure of \SI{700}{\K}, \SI{30}{\bar}, and for the stoichiometric
equivalence ratio. The rates of production of the species have been integrated until the time of
\SI{20}{\percent} fuel consumption. The results of this analysis are shown in \cref{tab:mv-radicals}
for the two models. The percentages shown in the \cref{tab:mv-radicals} are the percent of the fuel
consumed to form a particular fuel radical by all the reactions that can form that radical, and the
radicals are labeled according to the convention in \cref{fig:mv-structure}.

@@

At the relatively low temperature and high pressure condition of this analysis, all of the fuel is
consumed by \ce{H}-atom abstractions to form the fuel radicals shown. It can be seen that the two
models have quite different distributions of products from the first \ce{H}-abstraction reactions.
The model of \cite{Dievart2013} predicts that \ce{H}-abstraction from the second carbon is the most
prevalent, while the RMG model predicts that the radical on the fourth carbon in the chain will be
primarily formed. This is in line with the heats of formation in \cref{tab:heat-of-formation}, where
the most stable radical (i.e., the radical with the smallest heat of formation) is most likely to be
formed in each model.

To further compare the models with each other, the NASA polynomials representing the thermodynamic
properties of MV and the 5 fuel radicals from the model of \citet{Dievart2013} and are used to
replace the equivalent molecules in the RMG model. The results of a path analysis at the same
condition as the other analysis is shown in \cref{tab:mv-radicals} in the ``RMG switched'' column.
This analysis shows that the radical on the fourth carbon atom is still the most prevalant, despite
changing the heats of formation of the fuel and its radicals.

Taken together, these results show that the poor performance in a given model cannot be attributed
to a single source. There is a strong interaction between the thermodynamics of the species and the
kinetics of the reactions, requiring further detailed study of the methyl ester system to accurately
predict the low temperature ignition delays of methyl valerate.

>>>>

\bibliographystyle{elsarticle-num-CNF}
\usepackage{doi}
\usepackage[capitalize]{cleveref}
\begin{document}
%======================================================================
\begin{frontmatter}
\title{ Experiments and Modeling of the Autoignition of Methyl Pentanoate at Low to Intermediate Temperatures and Elevated Pressures in a Rapid Compression Machine }

@@

This work provides additional data for the autoignition of MV. Data is collected in a RCM under
engine relevant conditions spanning from \SIrange{15}{30}{\bar}, equivalence ratios \((\phi)\) from
\numrange{0.25}{2.0}, and temperatures from \SIrange{680}{1050}{\K}. The NTC region of MV is mapped
out to provide additional information on the fidelity of using MV as a biodiesel surrogate and on
the autoignition chemistry of large methyl esters.

@@

The RCM is equipped with heaters to control the initial temperature of the mixture. After filling in
the components to the mixing tanks, the heaters are switched on and the system is allowed
\SI{1.5}{\hour} to come to steady state. The mixing tanks are also equipped with magnetic stir bars
so the reactants are well mixed for the duration of the experiments. Previous work has shown this
procedure to completely vaporize the fuel and prevent fuel cracking during the heating process
\cite{Weber2011a,Kumar2009,Das2012}.

@@

The initial temperature is chosen such that the saturated vapor pressure \((P_{\text{sat}})\) of the
fuel at the initial temperature is at least twice the partial pressure of the fuel in the mixing
tank. The Antoine equation
%
\begin{equation}\label{eq:antoine}
    \logten{P_{\text{sat}}} = A - \frac{B}{T - C}
\end{equation}
%
is used to model the saturated vapor pressure of MV as a function of temperature \((T)\), where
\(A\), \(B\), and \(C\) are substance-specific coefficients, given in units of \si{\kelvin} and
\si{\kPa}. Coefficients for \cref{eq:antoine} are given in the literature by~\citet{Ortega2003},
\citet{Camacho2007}, and~\citet{Stephenson1987}. Unfortunately, the values of the coefficients are
different among all three references. Therefore, coefficients for use in \cref{eq:antoine} are
determined in this work by least squares fitting of the data of \citet{Ortega2003},
\citet{vanGenderen2002}, and \citet{Verevkin2008} using the \verb|curve_fit()| function of
SciPy~\cite{Jones2001} version 0.19.0. \Cref{fig:vapor-pressure} shows that the coefficients fitted
with this procedure give good agreement with the experimental data; values for the coefficients
computed in this work and in the literature works are given in \cref{tab:antoine}. The data and code
used to calculate the coefficients are provided in the Supplementary Material.

\begin{center}
    \captionsetup{type=table}
    \caption{Antoine Equation coefficients computed in this work and from the
    literature, in units of \si{\kelvin} and \si{\kPa}. The $2\sigma$ confidence interval is
    estimated by taking the square root of the diagonals of the covariance matrix returned from
    \texttt{curve\_fit()}}
    \label{tab:antoine}
    \begin{tabular}{cccccc}
        \toprule
        & $A$ & $B$ & $C$ & $T_{\text{min}}$, \si{\K} & $T_{\text{max}}$, \si{\K} \\
        \midrule
        This Work & 6.4030 & 1528.69 & 52.881 & 274.9 & 417.18 \\
        $2\sigma$ Confidence Interval & 0.0919 & 53.47 & 4.934 & --- & --- \\
        \citet{Ortega2003} & 6.23175 & 1429.00 & 62.30 & 364.75 & 417.18 \\
        \citet{Camacho2007} & 5.9644 & 1281.06 & 75.94 & 281 & 547 \\
        \citet{Stephenson1987} & 6.62646 & 1658.4 & 42.09 & 297 & 411 \\
        \bottomrule
    \end{tabular}
\end{center}

@@

\begin{center}
    \captionsetup{type=figure}
    \includegraphics[width=\textwidth]{figures/pressure-traces.pdf}
    \caption{Selected pressure traces around the NTC region of ignition delay.
    a) \(\phi=1.0,\ P_C=\SI{30}{\bar}\), b) \(\phi=2.0,\ P_C=\SI{15}{\bar}\).}
    \label{fig:pressure-traces}
\end{center}

@@

\begin{center}
    \captionsetup{type=figure}
    \resizebox{\textwidth}{!}{\input{figures/simulation-comparison.pgf}}
    \caption{Comparison of experimental and simulated ignition delays computed using the procedure
    described in \cref{sec:experimental-modeling}. a) \SI{15}{\bar}, b) \SI{30}{\bar}.}
    \label{fig:simulation-comparison}
\end{center}

At \SI{15}{\bar}, the experimental ignition delays are under-predicted by the \citet{Dievart2013}
model for the three equivalence ratios shown. For the \(\phi =\) \numlist{0.25;0.5} conditions, the
model appears to be predicting an NTC region of the ignition delays as the temperature decreases, as
judged by the increasing curvature of the simulations, although such a trend is not observed in the
data. However, at \(\phi = 2.0\), the model does not predict the presence of an NTC region, although
one is present in the experiments. Nonetheless, the agreement seems to be improving as the
temperature is decreased. Comparing the \citet{Dievart2013} model to the stoichiometric data at
\SI{30}{\bar}, we find a similar trend as the \(\phi=2.0,\ P_C=\SI{15}{\bar}\) data. The model does
not predict the NTC region found experimentally for the \(\phi = 1.0\) experiments, but the
agreement improves as the temperature decreases. Interestingly, two-stage ignition is predicted for
all of the \(\phi=1.0\) and \(\phi=2.0\) data shown in \cref{fig:simulation-comparison}. However,
the first-stage ignition delays are \SIrange{0.1}{0.5}{\ms} less than the overall ignition delays,
and are not shown on \cref{fig:simulation-comparison} because they are nearly indistinguishable from
the overall ignition delay.

\begin{center}
    \captionsetup{type=figure}
    \includegraphics[width=\textwidth]{figures/constant-volume-comparison.pdf}
    \caption{Comparison of simulated ignition delays computed in an adiabatic, constant volume
    system (dotted lines) and computed using the procedure described in
    \cref{sec:experimental-modeling} (solid lines). a) \SI{15}{\bar}, b) \SI{30}{\bar}.}
    \label{fig:conv-comparison}
\end{center}

To further understand the model of \citet{Dievart2013}, we have conducted adiabatic, constant volume
simulations (called CONV simulations), as these simulations are not linked to a particular
experiment by the volume trace and can be conducted over a wide range of temperatures.
\Cref{fig:conv-comparison} compares the CONV simulations to the RCM simulations shown in
\cref{fig:simulation-comparison}. From \cref{fig:conv-comparison}, it can be seen that the curvature
in the RCM simulations at $P_C = \SI{15}{\bar}$, $\phi =$ \numlist{0.25;0.5} is related to the NTC
region of the ignition delay, while the lack of curvature in the $P_C = \SI{15}{\bar}$, $\phi =
\num{2.0}$ and the $P_C = \SI{30}{\bar}$, $\phi = \num{1.0}$ simulations is because those lie on the
low-temperature side of the predicted NTC. It is clear from \cref{fig:conv-comparison} that the
model of \citet{Dievart2013} predicts the NTC to occur at too high a temperature relative to the
experiments.

To elucidate the underlying reasons for the disagreement between the \citet{Dievart2013} model and
the data, we constructed an additional model using RMG (see
\cref{sec:reaction-mechanism-generator}). As can be seen in \cref{fig:simulation-comparison}a, the
agreement between the RMG model (dashed lines) and the experimental data is similar to the
\citet{Dievart2013} model for the \SI{15}{\bar}, \(\phi =\) \numlist{0.25; 0.5} data. Moreover, the
RMG model predicts a similar NTC region as temperature is decreasing. For the \SI{15}{\bar}, \(\phi =
2.0\) data, the RMG model tends to over-predict the low-temperature ignition delays (i.e., those to
the right of the experimental NTC region on the Arrhenius plot), and does not predict the NTC region
found experimentally. As before, the trend at \SI{30}{\bar}, \(\phi=1.0\) is similar to the
\SI{15}{\bar}, \(\phi=2.0\) data; the RMG model over-predicts the low-temperature ignition delays
and does not predict the experimental NTC region. Finally, as in the \citet{Dievart2013} model,
two-stage ignition is predicted for all of the \(\phi=1.0\) and \(\phi=2.0\) data shown in
\cref{fig:simulation-comparison}. However, the first-stage ignition delays are
\SIrange{0.1}{0.5}{\ms} less than the overall ignition delays, and are not shown on
\cref{fig:simulation-comparison} because they are nearly indistinguishable from the overall ignition
delay.

@@

This difference in heats of formation affects the pathways that consume the fuel. By conducting a
reaction pathway analysis to determine which radicals are formed from the breakdown of the fuel, we
can analyze the proportion of each radical formed as the fuel breaks down during the autoignition
process. The following analysis is conducted for a constant volume, adiabatic simulation with
initial temperature and pressure of \SI{700}{\K}, \SI{30}{\bar}, and for the stoichiometric
equivalence ratio. The rates of production of the species have been integrated until the time of
\SI{20}{\percent} fuel consumption. The results of this analysis are shown in \cref{tab:mv-radicals}
for the two models. The percentages shown in \cref{tab:mv-radicals} are the percent of the fuel
consumed to form a particular fuel radical by all the reactions that can form that radical, and the
radicals are labeled according to the convention in \cref{fig:mv-structure}.

@@

At the relatively low temperature and high pressure condition of this analysis, all of the fuel is
consumed by \ce{H}-atom abstractions to form the fuel radicals shown. It can be seen that the two
models have quite different distributions of products from the first \ce{H}-abstraction reactions.
The model of \citet{Dievart2013} predicts that \ce{H}-abstraction from the second carbon is the most
prevalent, while the RMG model predicts that the radical on the fourth carbon in the chain will be
primarily formed. This is in line with the heats of formation in \cref{tab:heat-of-formation}, where
the most stable radical (i.e., the radical with the smallest heat of formation) is most likely to be
formed in each model.

To further compare the models with each other, the NASA polynomials representing the thermodynamic
properties of MV and the 5 fuel radicals from the model of \citet{Dievart2013} and are used to
replace the equivalent molecules in the RMG model. The results of a path analysis at the same
condition as the other analysis is shown in \cref{tab:mv-radicals} in the ``RMG switched'' column.
The results of the analysis of the ``RMG switched'' model show that the radical on the fourth carbon
is still the most prevalent, despite the heats of formation for the radicals in the ``RMG switched''
model being identical to the \citet{Dievart2013} model. This suggests that the reaction pathways
have a substantial impact on the simulation, in addition to the influence of the thermochemistry, as
discussed previously. Moreover, since the thermochemistry of the species in a reaction controls the
reverse reaction rate of a reaction, the RMG algorithm may miss important pathways due to improperly
estimated thermochemistry.

Taken together, these results show that the poor performance in a given model cannot be attributed
to a single source. Separating the influence of thermochemistry and kinetics requires further
detailed study of the methyl valerate system specifically, and methyl ester systems more generally.
Although such detailed work has begun, for example, with the work of \citet{Hayes2009}, further work
is required to accurately predict the low temperature ignition delays of methyl valerate.
